Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-17 latex/main.tex
Some details on the postprocessing

\subsection*{Data and code availability}

@@

\begin{itemize}
\item Snakemake workflow for aligning and converting the Viridian dataset to
VCF Zarr:
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
\end{itemize}

@@

@@

\subsubsection*{Reconstructing a pandemic-scale ARG}

%%% TODO: Much of this needs to be rewritten once the post-processing pipeline is finalised.
Using sc2ts, we reconstructed the ARG presented here in three stages:
(1) initialization, (2) extension, and (3) post-processing.

First, we initialized a base ARG.
We added the Wuhan-Hu-1 reference sample (MN908947.3) as the root node, dated December 8, 2019.
Then, we ran sc2ts unconditionally (i.e., without excluding samples by HMM cost) up to March 01, 2020 (inclusive).
This step is necessary to accumulate sufficient genetic diversity in the ARG
for subsequent sample matching and attachment in the next stage.

Second, we extended the base ARG by adding batches of samples collected on and after March 02, 2020.
Sample filtering, HMM cost-based rejection, and retrospective sample grouping, as described above,
are in full effect during this stage.

Finally, we performed the following post-processing steps
to modify certain non-topological properties of the ARG or
to enrich the ARG with additional information:
1. Moving mutations around recombinants? TODO
2. Adjusting recombination breakpoints. XX.
3. Mapping multi-base deletions. Multi-base deletions are challenging to handle, as they violate the Markov property of the HMM. In our ARG, insertions are deleted regions that acquire nucleotides (not necessarily the same nucleotides) without altering the genome length, because we discarded length-changing insertions when preprocessing the aligned sequences. We treated multi-base indels as multiple separate mutations in the HMM. We mapped them back onto the inferred ARG as single events.
4. Mapping mutations at masked sites. XX.
5. Adding exactly matched samples. During sample matching, samples identical to node(s) in a given ARG were not added to the ARG. Instead, after inference, we simply added the nodes to the ARG using the saved matching results. This allowed for computational savings, as it kept the number of possible parents available for matching to a minimum.
6. Estimating non-sample node ages. The ages of the inserted non-sample nodes (the nodes in the local trees and the recombination nodes) were initially assigned to arbitrarily determined values. We ran \textit{tsdate} (``variational gamma'', version XX; \cite{Wohns2022} Pope et al., 2025) on the ARG to adjust the ages of these nodes, while fixing the ages of the sample nodes to the reported collection times (defined as fractional years before the collection date of the last sample added to the ARG, which is February 20, 2025 in the ARG described in this study).
7. Assigning Pango lineages to non-sample nodes.

\subsubsection*{Post hoc parsimony mapping}
After the main inference step we mapped the variation data for 163 sites 
in the alignments to the ARG using the Fitch-Hartigan 
algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
implemented in tskit. In contrast to the main inference process where 
deletion characters are masked as missing data, we allow for 
5 different states in these parsimony calculations: A,C,G,T and -.
By doing this, we can approximate the mutation events causing
deletions, but with the clear caveat that each site is treated 
independently. 
Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
we chose 75 sites to remap

performed a large-scale analysis of indels
using a global GISAID dataset (sampled up to August, 2022).
We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
All of these indels are deletions and have high frequencies in the ARG
(sequence count divided by the total number of samples),
except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.

Based on the data of Table S4 of Li2023indels

>>>>

\subsection*{Data and code availability}
\label{sec:data_and_code_availability}

@@

[TODO fix the typesetting once all resources are compiled]

\begin{itemize}
\item Snakemake workflow for aligning and converting the Viridian dataset to
VCF Zarr:
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
\item Complete configuration for running sc2ts base ARG inference:
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}
\item Snakemake workflow for ARG post-processing
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/Snakefile}
\end{itemize}

@@

[TODO explain Fig S1 and S2]

@@

\subsubsection*{Primary ARG inference}
The ``base ARG'' we infer with sc2ts is the starting point for later analyses, 
and this step dominates overall processing time. To summarise, we first 
initialise the ARG using the Wuhan-Hu-1 reference sample (MN908947.3) as the
root node, and then ran the algorithm day-by-day using the parameters 
described in the preceding sections and documented in the configuration file 
(see \nameref{sec:data_and_code_availability}).
As discussed above,
computational resources are dominated by the LS model
(Figures~\ref{fig:samples_per_day},\ref{fig:inference_resources}) with all 
other aspects contributing negligibly.

\subsubsection*{ARG postprocessing}
The first step in ARG postprocessing is to examine putative recombinants to
determine whether a more parsimonious explanation is possible. This is done 
by rematching the inferred recombinant sequence against the ARG pruned back
to include only samples before the date of the recombinant. Additionally,
 each recombination node is examined to determine if any mutations 
are shared [TODO Yan write this up].
We refer to this process as ``long branch splitting''. 
% Data from
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
From this, 58 recombinants were identified as being more parsimoniously 
explained by matching to existing nodes in the ARG (inserted by 
subsequent parsimony improving heuristics) and 16 by long branch 
splitting. Of these 16 recombinants, two are noteworthy as 
being the origins of the Delta and BA.2 lineages in the ARG.
Both result in substantially fewer mutations (15 fewer for 
Delta, 6 for BA.2) than the previous recombinant HMM solution.
These major saltational lineages highlight the challenges 
of using the LS HMM with highly diverged samples.
We then updated the ARG to remove these 74 artefactual 
recombinants by ``rewiring'' the recombination node to the 
more parsimonious non-recombinant solution. This 
resulted in 16 additional nodes, 182 fewer mutations and reduced
the number of trees along the genome from 348 to 317.

% and requires a lot of off-topic explanation.
% It is also noteworthy that the origin of the BA.5
% lineage is identified as non-recombinant when long-branch
% splitting is applied to the daily ARG immediately after 
% its inclusion---for this reason, we have identif

The second postprocessing step performed is to add 1,252,208 samples
detected as exact matches during primary inference back
into the ARG. These are not included in the ARG during 
primary inference to reduce memory usage and improve 
LS HMM performance.

During primary inference, gap characters (``-'') are masked 
out as missing data, but this excludes both important information
about deletions from the ARG as well as potentially informative 
phylogenetic signal. 
We therefore performed post-hoc parsinony mapping at 163 sites to include 
the effects of deletions and to ensure that all Pango-informative
sites are included in the ARG.
Using deletions with a 
a minimum frequency of 1\% 
(count of sequences divided by 9,149,680 sequences analyzed)
from Table S1 of Li et al.\cite{Li2023indels},
we chose 75 sites to remap.
In addition, we remapped data at 88 sites that were excluded
from primary inference and are defined as Pango lineage 
defining sites by
 Freyja\cite{Karthikeyan2022}
(\url{https://raw.githubusercontent.com/andersen-lab/Freyja/refs/heads/main/freyja/data/lineage\_mutations.json}).
Alignment data for these 163 sites was then mapped to the 
ARG using the Fitch-Hartigan 
algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
implemented in tskit, including the gap character as a state 
along with the 4 nucleotides.
Following this, we iteratively applied the node parsimony 
heuristics in sc2ts until there were no further reductions in 
the numbers of mutations.

% Shifting breakpoints is a minor point, not worth mentioning?
We computed Pango assignments for all 2,747,985 nodes in the ARG by first 
exporting their aligned sequences to a 77GiB FASTA file and 
then running pangolin (version 4.3.1; pangolin-data version 1.29).
The resulting Pango and Scorpio designations were then stored as 
metadata associated with each node in the ARG.

When then estimated non-sample node dates using 
the ``variational-gamma'' algorithm in tsdate\cite{Wohns2022}
version 0.2.3. [TODO Say something above the node ages?]

Finally, we reduced the metadata stored in the ARG to a minimal
set (node sample IDs, along with computed Pango and Scorpio 
designations), encoded to enable efficient columnar access,
and compressed the file using tszip (version 0.2.5). The final 
file requires 32MiB of storage.

This full post-processing pipeline is encoded as a reproducible Snakemake
workflow (see \nameref{sec:data_and_code_availability}).
